### 2019-10-12 main_text/main_text.tex
Edits to model section

A key element of this mechanism is the presence of a transcription factor
which is always expressed at a low level (often at copy numbers
of order $\sim10$) and which binds to the transporter substrate.
This is resource efficient for the cell, as this low copy number
transcription factor acts as an ``always on'' sensor to detect
the substrate, allowing high copy numbers of the membrane transporter
and its attendant operon to be expressed only when
their substrate is actually present. The transcription factors
LacI, AraC, and XylR all appear to fill this
role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
\todo{do we have a cite for xylR too?}

@@

\section*{Model}
\subsection*{Step by step modeling of the system}
In this section, we present our model of the \emph{xapAB} genetic circuit.
Fig~\ref{fig2:model} shows an overview of this model. It consists of the
following steps: 
\begin{itemize}
	\item Small amounts of xanthosine are transported into and out of cell in various ways (discussed in more detail later on).
	\item XapR is induced by xanthosine.
	\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
	\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
	\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
	\item At some point, the proteins decay.
\end{itemize}
Because xanthosine induces the transcription factor XapR, we have a positive
and negative feedback loop due to XapB and XapA, respectively. The remainder
of this subsection discusses each of the above steps in detail, leading us
to a set of two coupled ODE's. More in-depth explanations can be found
in~\nameref{S1_Text}.

\begin{figure}[h!]
	\centering
	\includegraphics{media/XapSystem.eps}
	\caption{{\bf Model of the \emph{xapAB} circuit.}
		The XapR dimers are induced by xanthosine and the induced XapR binds
		cooperatively as an activator to the \emph{xapAB} promoter. For
		these two steps, quasi-equilibrium is assumed. If both XapR binding
		sites are occupied and the polymerase is bound, the gene is
		transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
		$\gamma_{\n{m}}$, and both proteins are translated at rate
		$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
		xanthosine with Michaelis-Menten parameters $k_{a}$ and $K_a$.
		Similarly, XapB imports and exports xanthosine with parameters
		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
		respectively. Furthermore, xanthosine enters and leaves the cell in
		other ways, proportional to rates $k_{\n{nup}}$ and $\xi
		k_{\n{nup}}$, respectively.}
	\label{fig2:model}
\end{figure}

\paragraph*{Induction of XapR.} 
We treat dimers as the only form of XapR that appears in the cell. Each
dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
model is used to describe the fraction of XapR dimers in the active state: 
\begin{eqnarray}
\label{eq:MWC}
\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}
\end{eqnarray}
Here, $\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
respectively. Furthermore, $K_{\n{xI}}$ and $K_{\n{xA}}$ are the
dissociation constants of xanthosine to the inactive and the active XapR
dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
between the inactive and the active state. We expect $\Delta E_{\n{x}} > 0$
and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
XapR being mainly inactive in the absence of xanthosine and becoming mostly
active at high concentrations of xanthosine. A detailed discussion of the
MWC model can be found in~\cite{Marzen2013}.

@@

In Fig~\ref{fig3:states}, all possible states of the promoter in our model
and their corresponding thermodynamic weights are shown. There, $\n{[P]}$
denotes the polymerase concentration and $\Delta E_{\n{coop}}$ stands for
the interaction energy of the two XapR dimers. If cooperativity in
transcription factor binding is neglected, this is set to zero. Furthermore,
$K_{\n{XapR}}$ and $K_{\n{P}}$ denote the dissociation constant of XapR and
polymerase to the promoter, respectively. In statistical mechanics language
this is equivalent to $\frac{N_{\mathrm{NS}}}{V} e^{\beta \Delta E_{i}}$
with $N_{\mathrm{NS}}$ being the number of non-specific binding sites on the
DNA, $V$ the volume of the cell, and $\Delta E_{i}$ the interaction energy
of $i=$XapR or $i=$polymerase with the promoter.

@@

We consider only the state where both XapR binding sites are occupied as
active and all other states as inactive, meaning they have transcription
rate $0$. Experiments show that the expression becomes very weak when one of
the XapR binding sites is removed from the promoter, suggesting that this
simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
the bistable parameter range, considering the single occupancy states as
active instead has almost no influence on the results (see
also~\nameref{S1_Text}).

With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
probability of the promoter being in the active state, we obtain
\begin{eqnarray}
\label{eq:mbasic}
\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
\frac{
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
	e^{- \beta \Delta E_{\n{coop}}}
}{
	1 + 
	2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
}
\end{eqnarray}
Here, $w_i$ stands for the thermodynamic weight of the ith state in the
order in which they are listed in Fig~\ref{fig3:states}. As written above,
the partition function factorizes into a polymerase and a XapR term because
of our assumption of independent binding, which is further discussed
in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
p_{\n{active}}$, which we will need in the next paragraph.
In~\nameref{S1_Text}, the full chemical master equation is shown, which
leads to a Poisson distribution of the mRNA concentration with the same
mean.

>>>>

A key element of this mechanism is the presence of a transcription factor
which is always expressed at a low level (often at copy numbers
of order $\sim10$) and which binds to the transporter substrate.
This is resource efficient for the cell, as this low copy number
transcription factor acts as an ``always on'' sensor to detect
the substrate, allowing high copy numbers of the membrane transporter
and its attendant operon to be expressed only when
their substrate is actually present. The transcription factors
LacI, AraC, and XylR all appear to fill this
role~\cite{Novick1957,Narang2008,Choi2008,Ozbudak2004,Fritz2014,Jenkins2017,Siegele1997}.
\todo{MJM: do we have a cite for xylR too?}

@@

\section*{Model}
\subsection*{Step by step modeling of the system}
In this section, we present our model of the \emph{xapAB} genetic circuit.
Fig~\ref{fig2:model} shows an overview of this model.
The qualitative picture of the circuit switching its state is as follows:
\begin{itemize}
	\item In the initial absense of XapB, small amounts of xanthosine are
		nonspecifically transported into and out of the cell (discussed in more detail below).
	\todo{MJM: ``nonspecific transport'' is a bit misleading:
	it's actually weakly specific transporters, but the precise mechanism is unimportant
	\& I think it better to emphasize generality of the model. Does this seem fine?}
	\item XapR is induced by xanthosine.
	\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
	\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
	\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
	\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
\end{itemize}
Because xanthosine induces the transcription factor XapR, we have positive
and negative feedback loops due to XapB and XapA, respectively. The remainder
of this subsection discusses each of the above steps in detail, leading us
to a set of two coupled ODEs. More in-depth explanations can be found
in~\nameref{S1_Text}.

\begin{figure}[h!]
	\centering
	\includegraphics{media/XapSystem.eps}
	\caption{{\bf Model of the \emph{xapAB} circuit.}
		The XapR dimers are induced by xanthosine and the induced XapR binds
		cooperatively as an activator to the \emph{xapAB} promoter. For
		these two steps, quasi-equilibrium is assumed. If both XapR binding
		sites are occupied and the polymerase is bound, the gene is
		transcribed at rate $r_{\n{m}}$. The mRNA decays at rate
		$\gamma_{\n{m}}$, and both proteins are translated at rate
		$r_{\n{p}}$ and decay at rate $\gamma_{\n{p}}$. XapA degrades
		xanthosine with Michaelis-Menten parameters $k_{a}$ and $K_a$.
		Similarly, XapB imports and exports xanthosine with parameters
		$k_{\n{b,i}}, K_{\n{b,i}}$ and $k_{\n{b,e}}, K_{\n{b,e}}$,
		respectively. Furthermore, xanthosine enters and leaves the cell
		through non-specific transport, proportional to rates $k_{\n{nup}}$ and $\xi
		k_{\n{nup}}$, respectively.}
	\label{fig2:model}
\end{figure}

\paragraph*{Induction of XapR.} 
We treat dimers as the only form of XapR that appears in the cell. Each
dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
model is used to describe the fraction of XapR dimers in the active state,
which has the form
\begin{eqnarray}
\label{eq:MWC}
\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}.
\end{eqnarray}
Here, $\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
respectively. Furthermore, $K_{\n{xI}}$ and $K_{\n{xA}}$ are the
dissociation constants of xanthosine to the inactive and the active XapR
dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
between the inactive and the active state. We expect $\Delta E_{\n{x}} > 0$
and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
XapR being mainly inactive in the absence of xanthosine and becoming mostly
active at high concentrations of xanthosine. A detailed discussion of the
MWC model can be found in~\cite{Marzen2013}.

@@

In Fig~\ref{fig3:states}, all possible states of the promoter in our model
and their corresponding thermodynamic weights are shown. $\n{[P]}$
denotes the polymerase concentration, and $\Delta E_{\n{coop}}$ stands for
the interaction energy of the two XapR dimers. If cooperativity in
transcription factor binding is neglected, this is set to zero. Furthermore,
$K_{\n{XapR}}$ and $K_{\n{P}}$ denote the dissociation constant of XapR and
polymerase to the promoter, respectively. In statistical mechanics language
these dissociation constants are equivalent to
$\frac{N_{\n{NS}}}{V} e^{\beta \Delta E_{\n{XapR}}}$
and $\frac{N_{\n{NS}}}{V} e^{\beta \Delta E_{\n{P}}}$, respectively,
with $N_{\mathrm{NS}}$ being the number of non-specific binding sites on the
DNA, $V$ the volume of the cell, and $\Delta E_{\n{XapR}}$ and
$\Delta E_{\n{XapR}}$, respectively, the interaction energies
of XapR or polymerase with the promoter.

@@

We consider only the state where both XapR binding sites are occupied as
active and all other states as inactive, meaning they have transcription
rate equal zero. Experiments show that the expression becomes very weak when one of
the XapR binding sites is removed from the promoter, suggesting that this
simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
the bistable parameter range, considering the single occupancy states as
active instead has almost no influence on the results (see
also~\nameref{S1_Text}).

With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
probability of the promoter being in the active state, we obtain
\begin{eqnarray}
\label{eq:mbasic}
\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
\frac{
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
	e^{- \beta \Delta E_{\n{coop}}}
}{
	1 + 
	2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
	\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
}
\end{eqnarray}
Here, $w_i$ stands for the thermodynamic weight of the ith state in the
order in which they are listed in Fig~\ref{fig3:states}. As written above,
the partition function factorizes into a polymerase and a XapR term because
of our assumption of independent binding, which is further discussed
in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
p_{\n{active}}$, which we will need in the next paragraph.
In~\nameref{S1_Text}, the full chemical master equation is shown, which
leads to a \todo{MJM: I worry referees will give us grief: ``Real dists aren't Poisson,'' even though it's irrelevant}
Poisson distribution of the mRNA concentration with the same
mean.
